Computing properties of independent sets by generic programming tensor networks \thanks{\funding{...}}

### 2021-12-23 paper/paper.tex
update

\begin{abstract}
We introduce a new approach to compute various properties of independent sets by encoding the problems and graphs into tensor networks with different tensor element algebra.
The type of graph problems that can be computed using this method includes: independence number, the number of independent sets, independence polynomial, maximal independence polynomial, and both counting and enumeration of all maximum independence sets, the large-size independent sets, and all maximal independent sets.
Although the computational complexity inevitably scales exponentially with treewidth of the graph, our method using tensor network contraction with generic programming is highly versatile and is able to compute properties that are otherwise not known to be possible for up to several hundred vertices.
At the same time, our algorithms are very simple to implement and one can directly utilize recent advances in tensor network contraction techniques.
To demonstrate the versatility of this tool, we apply it to a few examples including the calculations of the hard-square entropy constant, Euler characteristics of the independence complex, partition functions, and finite-temperature phase transitions on square-lattice graphs.
\end{abstract}

@@

\section{Introduction}
\blue{overlap gap property~\cite{Gamarnik2013, Gamarnik2019}, independence polynomial at -1~\cite{Bousquet2008,Levit2009},
two motivations of computing independence polynomial: 1. Lee-Yang zero~\cite{Lee1952,Yang1952}, 2. Lov{\'a}sz local lemma~\cite{Scott2005}.
}
In graph theory and combinatorics, there are many interesting but hard computational problems concerning properties of independent sets.
An independent set is a set of vertices in a graph where no two vertices are adjacent to each other.
More formally, given an undirected graph $G = (V,E)$, an independent set $I \subseteq V$ is a set that for any $u,v \in I$, there is no edge connecting $u$ and $v$ in $G$.
The problem of finding an independent set of the largest possible size, the maximum independent set (MIS), is a paradigmatic NP-complete problem~\cite{Hastad1996}.
The size of an MIS of a graph $G$ is called the independence number, denoted as $\alpha(G) \equiv \max_{I}|I|$. 
Finding $\alpha(G)$ is a hard computational problem; moreover, it is NP-hard to even approximate $\alpha(G)$ within a factor $|V|^{1-\epsilon}$ for any positive $\epsilon$.
Naive exhaustive search for the MIS requires time $O\left(2^{|V|} \right)$. More efficient exact algorithms include the branching algorithms~\cite{Tarjan1977,Robson1986} and dynamic programming.
Sophisticated branching algorithms can reduce the base of the exponential time scaling to, e.g., $1.1996^{|V|} {|V|}^{O(1)}$~\cite{Xiao2017}.
Dynamic programming approaches~\cite{Courcelle1990, Fomin2013} works better for graphs with a small treewidth $\text{tw}(G)$;
these methods can produce algorithms of complexity $O(2^{\text{tw}(G)} \text{tw}(G) |V|)$.

The independent set problem is closely related to the clique problem and the vertex cover problem~\cite{Moore2011};
more concretely, an maximal independent set of a graph $G$ corresponds to a maximum clique in the complement graph of $G$,
and the complement vertex set of an MIS corresponds to a minimum vertex cover of the graph $G$.
In this paper, the term ``maximal'' has a different meaning from ``maximum''  which we discussed above; a maximal independent set is an independent set that is not a subset of any other independent set, but its size may not be the maximum.
Together, these problems find a wide range of applications in scheduling, logistics, social network analysis, bioinformatics, wireless networks and telecommunication,
map labelling, computer vision, etc.~\cite{Butenko2003, Wu2015}.
Besides the standard problem of finding the independence number and an MIS, there are many other interesting problems pertaining to independent sets, such as the number of independent sets,
independence polynomial, maximal independence polynomial, and enumeration of MIS or maximal independent sets.
Some of these problems are of great interest in physics applications such as the hard-core lattice gas model~\cite{Dyre2016, Fernandes2007}
in statistical mechanics and the Rydberg hamiltonian with neutral atoms~\cite{Pichler2018} \red{cite experiment when it's ready};
they can, for example, be used to understand phase transitions and to identify harder graphs in an ensemble of graphs \red{cite experiment}.

In this work, we introduce a tensor-network based framework to compute the various properties pertaining to independent sets.
We map the independent set problem into generic tensor network contraction with specially designed tensor element algebra.
Our approach does not necessarily provide a better time complexity compared to dynamic programming considering the fact that the tensor network contraction can be regarded as a type of concrete implementation of dynamic programming,
but the tensor network approach is highly versatile so different problems can be generically solved with minimal implementation efforts without loss of efficiency.
We benchmark our algorithms by computing various independent set properties on CPUs and GPUs, the implementation of which benefits from recent advances in tensor-network contraction for the purposes of quantum circuit simulations~\cite{Gray2021, Pan2021, Kalachev2021}.
Lastly, we provide a few examples and show that the toolbox can be used to calculate the hard-square entropy constant,
Euler characteristics of the independence complex, partition functions, and finite-temperature phase transitions on square-lattice graphs.

>>>>

\begin{abstract}
We introduce a method of using generic programming tensor network to compute various properties of independent sets.
This method is versertile. By engineering tensor element algebra, the type of graph problems that can be computed using this method includes: independence number, the number of independent sets, independence polynomial, maximal independence polynomial, and both counting and enumeration of all maximum independence sets, the large-size independent sets, and all maximal independent sets.
%Although the computational complexity inevitably scales exponentially with treewidth of the graph,
%our method using tensor network contraction with generic programming is highly versatile and is able to compute properties that are otherwise not known to be possible for up to several hundred vertices.
At the same time, our algorithms are very simple to implement and one can directly utilize recent advances in tensor network contraction techniques like near optimal contraction order finding and slicing.
We can easily contract a tensor network with thousands of tensors with an algorithmic complexity $2^{{\rm tw}(G)}$, with ${\rm tw} (G)$ the tree width of the problem graph.
To demonstrate the versatility of this tool, we apply it to a few examples including the calculations of the hard-square entropy constant, Euler characteristics of the independence complex, partition functions, and finite-temperature phase transitions on square-lattice graphs.
\end{abstract}

@@

\section{Introduction}
\blue{independence polynomial at -1~\cite{Bousquet2008,Levit2009},
two motivations of computing independence polynomial: 1. Lee-Yang zero~\cite{Lee1952,Yang1952}, 2. Lov{\'a}sz local lemma~\cite{Scott2005}.
}
In graph theory and combinatorial optimization, there are many interesting hard computational problems concerning properties of independent sets.
%An independent set of a graph is a subset of its vertices that no two vertices in this set are adjacent to each other.
For an undirected graph $G = (V,E)$, an independent set $I \subseteq V$ is a set of vertices that for any $u,v \in I$, $(u, v) \not\in E$. %there is no edge connecting $u$ and $v$ in $G$.
The primary insterest of many computational complexity theorists is finding the maximum size of such vertex sets be $\alpha(G) \equiv \max_{I}|I|$ and one of the sets having this size.
Finding $\alpha(G)$ is a hard computational problem; moreover, it is NP-hard to even approximate $\alpha(G)$ within a factor $|V|^{1-\epsilon}$ for an arbitrarily small positive $\epsilon$ ~\cite{Hastad1996}.
Naive exhaustive search for an MIS requires computing time $O\left(2^{|V|} \right)$.
People developed more efficient exact algorithms for finding MISs, including the branching algorithms~\cite{Tarjan1977,Robson1986} and dynamic programming.
The branching algorithms can reduce the base of the exponential time scaling to, e.g., $1.1996^{|V|} {|V|}^{O(1)}$~\cite{Xiao2017}, 
while dynamic programming approaches~\cite{Courcelle1990, Fomin2013} works better for graphs with a small treewidth $\text{tw}(G)$;
these methods can produce algorithms of complexity $O(2^{\text{tw}(G)} \text{tw}(G) |V|)$.
Although it is a hard problem, people keep chanllenging themself to find a better algorithm for solving the MIS problem;
not only for it has a wide range of applications in scheduling, logistics, social network analysis, bioinformatics, wireless networks and telecommunication,
map labelling, computer vision, etc.~\cite{Butenko2003, Wu2015}, but also for it is a well know NP-complete problem that can be mapped from
 to many other important combinatorial optimization problems in polynomial time such as circuit satisfiability~\cite{}, the maximum clique problem and the maximum vertex cover~\cite{Moore2011} problems.

%more concretely, an maximal independent set of a graph $G$ corresponds to a maximum clique in the complement graph of $G$,
%and the complement vertex set of an MIS corresponds to a minimum vertex cover of the graph $G$.
%In this paper, the term ``maximal'' has a different meaning from ``maximum''  which we discussed above; a maximal independent set is an independent set that is not a subset of any other independent set, but its size may not be the maximum.
In this paper, we do not limit our discussion the finding one of the maximum indepedent sets and its size.
There are many other interesting problems pertaining to independent sets, such as the number of independent sets,
independence polynomial, maximal independence polynomial, and enumeration of MIS or maximal independent sets.
Some of these problems are of great interest in physics applications such as the hard-core lattice gas model~\cite{Dyre2016, Fernandes2007}
in statistical mechanics and the Rydberg hamiltonian with neutral atoms~\cite{Pichler2018} \red{cite experiment when it's ready};
they can, for example, be used to understand phase transitions, to analyse the overlap gap properties~\cite{Gamarnik2013, Gamarnik2019} and to identify harder graphs in an ensemble of graphs \red{cite experiment}.

In this work, we introduce a tensor-network based framework to compute the various properties pertaining to independent sets.
We map the independent set problem into generic tensor network contraction with specially designed tensor element algebra.
Its algorithmic complexity is similar to that of dynamic programming, i.e. scales exponentially as the tree width of the graph ${\rm tw}(G)$.
%But the tensor network approach is highly versatile so different problems can be generically solved with minimal implementation efforts without loss of efficiency.
We benchmark our algorithms by computing various independent set properties on CPUs and GPUs.
Its high performance benefits from recent advances in random tensor-network contraction for the purposes of quantum circuit simulations~\cite{Gray2021, Pan2021, Kalachev2021}.
People developed algorithms to optimize the contraction order for tensor networks with thousands of vertices,
while the optimized contraction order gives us a near optimal space complexity $2^{\sim{\rm tw}(G)}$ to contract the tensor network.
For cases when the tensors being too large to fit into a GPU, slicing technique can be used to further reduce the space consumption.
The implementation of our algorithm is publicly available online: ~\cite{repolink}.
Lastly, we provide a few examples and show that this toolbox can be used to calculate the hard-square entropy constant,
Euler characteristics of the independence complex, partition functions, and finite-temperature phase transitions on square-lattice graphs.
